---
title: "Getting started with boldR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with boldR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
```

> **fmriprep gets you clean data. boldR gets you useful data.**

`boldR` picks up exactly where [fMRIPrep](https://fmriprep.org) leaves off.
It reads preprocessed BIDS derivatives and transforms them into analysis-ready
outputs: ROI timeseries, functional connectivity matrices, and voxelwise
quality metrics — all exportable to `syncR` for integration with sleep and
circadian data.

---

## Installation

```{r install}
# install.packages("pak")
pak::pak("circadia-bio/boldR")
```

---

## Basic workflow

### 1. Read fMRIPrep derivatives

```{r read}
library(boldR)

fprep <- read_fmriprep(
  derivatives_dir = "data/derivatives/fmriprep",
  subject         = "01",
  task            = "rest",
  space           = "MNI152NLin2009cAsym"
)
fprep
```

### 2. Prepare the BOLD object

```{r prepare}
bold <- prepare_bold(
  fmriprep    = fprep,
  tr          = 2,      # seconds; or NULL to read from NIfTI header
  drop_volumes = 4      # discard first 4 dummy scans
)
bold
```

### 3. Load an atlas and parcellate

`boldR` ships with several built-in atlases and accepts any custom NIfTI
label image via the `boldR_atlas` schema.

```{r atlas}
# See available built-ins
list_atlases()

# Load Schaefer 100-parcel atlas
atlas <- load_atlas("schaefer100_7n")

# Extract mean timeseries per ROI
parcel <- parcellate(bold, atlas)
dim(parcel$timeseries)  # timepoints × 100
```

#### Custom atlas

```{r custom-atlas}
custom <- load_atlas(
  atlas  = "my_parcellation.nii.gz",
  labels = paste0("Region_", 1:80),
  space  = "MNI152NLin2009cAsym"
)
parcel_custom <- parcellate(bold, custom)
```

### 4. Compute metrics

```{r metrics}
# Functional connectivity matrix (Fisher z)
fc <- compute_fc(parcel)
dim(fc$z)  # 100 × 100

# ROI summary statistics
roi_stats <- compute_roi_metrics(parcel)
roi_stats

# Voxelwise temporal SNR
tsnr <- compute_tsnr(bold)
tsnr$median_tsnr
```

### 5. Export for syncR

```{r export}
bold_export <- export_bold(parcel, fc = fc)
bold_export

# In your syncR pipeline:
# syncR::sync(
#   zeitR    = actigraphy_metrics,
#   slumbR   = sleep_diary,
#   tallieR  = questionnaires,
#   bold     = bold_export
# )
```

---

## Ecosystem

`boldR` is part of the Circadia Lab R ecosystem. See
[circadia-lab.uk](https://circadia-lab.uk) for the full package suite.
